Estimating the number of clusters using Cross Validation

### 2015-06-18 cvclust.tex
finish proof of kmeans existence

\begin{lemma}

@@

Pick~$r$ so that the ball~$K$ of radius~$r$ centered at the origin has
positive $P$~measure.  Take $M$ to be large enough such that
\(
  (M - r)^2 P(K) > \int \|x\|^2 P(dx).
\)
Let $\sC_1$ denote the class of sets $A$ with at most $k$ points and at least
one point in in $B(M)$.  Let $A_0$ be the singleton set containing the origin.
If no point of $A$ is in $B(M)$, then
\(
  W(A, P) \geq (M - r)^2 P(K) > W(A_0, P).
\)
Put concisely, if $A \in \sA_k \setminus \sC_1$, then
\(
  W(A, P) > W(A_0, P) \geq m_k(P).
\)

@@

We first show that if $M$ is large enough, then there exists a set $A \in \sE_k$
and a positive $\varepsilon$ with $\varepsilon + W(A,P) < m_{k-1}(P)$.
Since the support of $P$ contains least $k$ points, there exists an $x_0$ not
in $\bar A_{k-1}$ that lies in the support of $P$.  Ensure that $M$ is large
enough so that $x_0 \in B(5M)$.  Take $A \in \sE_k$ to be the set of $k$
points gotten by adjoining $x_0$ to $\bar A_{k-1}$.  Let
\(
  \delta = \min_{a \in \bar A_{k-1}} \| x_0 - a \|
\)
and set $R = \{ x \in \mathbb{R}^d : \|x - x_0\| < \delta/3 \}$; since $x_0$
is in the support of $P$, the neighborhood $R$ has positive $P$~measure.
Further,
\[
  \int_{R} \min_{a \in \bar A_{k-1}} \|x - a\|^2 P(dx)
  \geq
  \int_{R} \min_{a \in \bar A_{k-1}} (\|x_0 - a\| - \delta/3)^2 P(dx)
  = (4/9) \delta^2 P(R),
\]
and
\[
  \int_{R} \min_{a \in A} \|x - a\|^2 P(dx)
  =
  \int_{R} \|x - x_0\|^2 P(dx)
  \leq
  (1/9) \delta^2 P(R).
\]
Thus, if
\(
  \varepsilon < (3/9) \delta^2 P(R)
\)
then
\(
  m_{k-1}(P) - W(A, P) = 
  W(\bar A_{k-1}, P) - W(A, P) > \varepsilon.
\)

>>>>

\begin{lemma}\label{lem:kmeans-existence}

@@

Pick~$r$ so that the ball~$K$ of radius~$r$ centered at the origin has
positive $P$~measure.  Take $M$ to be large enough such that
\(
  (M - r)^2 P(K) > \int \|x\|^2 P(dx).
\)
Let $\sC_k$ denote the class of sets $A$ with at most $k$ points and at least
one point in in $B(M)$.  Let $A_0$ be the singleton set containing the origin.
If no point of $A$ is in $B(M)$, then
\(
  W(A, P) \geq (M - r)^2 P(K) > W(A_0, P).
\)
Put concisely, if $A \in \sA_k \setminus \sC_k$, then
\(
  W(A, P) > W(A_0, P) \geq m_k(P).
\)

@@

We first show that if $M$ is large enough, then there exists a set $B \in \sE_k$
and a positive $\varepsilon$ with $W(B,P) < m_{k-1}(P) - \varepsilon$.  This
will establish that $m_{k}(P) < m_{k-1}(P)$, and the set $B$ will have further
utility later in the proof.
Take $\bar A_{k-1} \in \sE_{k-1}$ to be a set with at most $k-1$ points satisfying
$m_{k-1}(P) = W(\bar A_{k-1},P)$.
Since the support of $P$ contains least $k$ points, there exists a point~$b_1$ not
in $\bar A_{k-1}$ that lies in the support of $P$.  Ensure that $M$ is large
enough so that $b_1 \in B(5M)$.  Take $B \in \sE_k$ to be the set of $k$
points gotten by adjoining $b_1$ to $\bar A_{k-1}$.  Let
\(
  \delta = \min_{a \in \bar A_{k-1}} \| b_1 - a \|
\)
and set $R = \{ x \in \mathbb{R}^d : \|x - b_1\| < \delta/3 \}$; since $b_1$
is in the support of $P$, the neighborhood $R$ has positive $P$~measure.
Further,
\[
  \int_{R} \min_{a \in \bar A_{k-1}} \|x - a\|^2 P(dx)
  \geq
  \int_{R} \min_{a \in \bar A_{k-1}} (\|b_1 - a\| - \delta/3)^2 P(dx)
  = (4/9) \delta^2 P(R),
\]
and
\[
  \int_{R} \min_{b \in B} \|x - b\|^2 P(dx)
  =
  \int_{R} \|x - b_1\|^2 P(dx)
  \leq
  (1/9) \delta^2 P(R).
\]
Thus, if
\(
  \varepsilon < (3/9) \delta^2 P(R)
\)
then
\(
  W(B,P) \leq W(\bar A_{k-1},P) - (3/9) \delta^2 P(R)
         < m_{k-1}(P) - \varepsilon.
\)

Now, ensure that $M$ large enough so that
\[
  4 \int_{\|x\| \geq 2M} \|x\|^2 P(dx) < \varepsilon.
\]
We will now show that if $A \in \sC_k \setminus \sE_k$, then
$W(A,P) > W(B,P) \geq m_{k}(P)$, which ensures that $W(\cdot, P)$ does not
attain its infimum over $\sA_k$ outside of $\sE_k$.  To see that this holds,
suppose that $A$ is any set $\sC_k \setminus \sE_k$.  This set has at least
one point, $a_1$, in $B(M)$, and at least one point outside $B(5M)$.  Consider
thet set $A^\ast$ formed by removing all points of $A$ outside $B(5M)$.  Any
point $x$ closer to a cluster center outisde $B(5M)$ than to $a_1$ must
satisfy $\|x\| \geq 2M$.  Thus,
\begin{align*}
  W(A^\ast, P)
    &\leq W(A,P) + \int_{\|x\| \geq 2M} \|x - a_1\|^2 P(dx) \\
    &\leq W(A,P) + \int_{\|x\| \geq 2M} (\|x\| + \|a_1\|)^2 P(dx) \\
    &\leq W(A,P) + 4 \int_{\|x\| \geq 2M} \|x\|^2 P(dx) \\
    &< W(A,P) + \varepsilon.
\end{align*}
Since $A^\ast$ has $k-1$ or fewer points, it must satisfy
$W(A^\ast, P) \geq m_{k-1}(P)$, so that
$W(A, P) > m_{k-1}(P) - \varepsilon > W(B, P)$.

We have now established that $m_k(P) = \inf \{ W(A,P) : A \in \sE_k \}$,
and that $W(\cdot, P)$ does not obtain its infimum over $\sA_k$ outside of
$\sE_k$.  The final step of the proof is to show $W(\cdot, P)$ obtains its
infimum over $\sE_k$.  This follows from the extreme value theorem, since
$\sE_k$ is compact and, by Lemma~\ref{lem:pollard-continuous}, the map $A
\mapsto W(A,P)$ is continuous over $\sA_k$.
